---
title: "2_EDA"
output: pdf_document
date: "2025-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Cleaning and Exploratory Data Analysis

This R markdown document will be used to perform Exploratory Data Analysis (EDA) on the merged_df_raw that was developed in the previous workbook `1_data_extraction.Rmd`. 

## Data Importing

```{r libraries}
library(tidyverse)
library(readr)
library(ggplot2)
library(corrplot)
library(caret)
library(here)
library(naniar)
```


```{r data_import}
merged_df <- read_csv('../Data/Raw/merged_df_raw.csv')
head(merged_df)
```
## Categorical Conversions

```{r}
## categorical features and their levels and labels
#sex
merged_df$sex <- factor(merged_df$sex, 
                    levels = c(0, 1), 
                    labels = c("Female", "Male")
                    )
#painloc
#unique(merged_df$painloc)
merged_df$painloc <- factor(merged_df$painloc,
                        levels = c(0, 1), 
                        labels = c("Otherwise", "Substernal")
                        )
#painexer
#unique(merged_df$painexer)
merged_df$painexer <- factor(merged_df$painexer,
                        levels = c(0, 1), 
                        labels = c("Otherwise", "Provoked by Exertion")
                        )
#relrest
#unique(merged_df$relrest)
merged_df$relrest <- factor(merged_df$relrest,
                        levels = c(0, 1), 
                        labels = c("Otherwise", "Relieved after rest")
                        )
#cp
#unique(merged_df$cp)
merged_df$cp <- factor(merged_df$cp,
                        levels = c(1, 2, 3, 4), 
                        labels = c("Typical Angina", "Atypical Angina", 
                                   "Non-anginal Pain", "Asymptomatic"), 
                   ordered = TRUE)
#smoke
#unique(merged_df$smoke)
merged_df$smoke <- factor(merged_df$smoke,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#fbs
#unique(merged_df$fbs)
merged_df$fbs <- factor(merged_df$fbs,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#dm
#unique(merged_df$dm)
merged_df$dm <- factor(merged_df$dm,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#famhist
#unique(merged_df$famhist)
merged_df$famhist <- factor(merged_df$famhist,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#restecg
#unique(merged_df$restecg)
merged_df$restecg <- factor(merged_df$restecg,
                        levels = c(0, 1, 2), 
                        labels = c("Normal", "Abnormal", "Hypertrophy"), 
                   ordered = TRUE)
#dig
#unique(merged_df$dig)
merged_df$dig <- factor(merged_df$dig,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#prop
#unique(merged_df$prop)
# find the number of 22's a feature outside the range/
#sum(merged_df$prop == 22, na.rm = TRUE)
# convert the 22 to NA
merged_df <- merged_df %>%
  mutate(prop = ifelse(prop == 22, NA, prop))
#create factor
merged_df$prop <- factor(merged_df$prop,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#nitr
#unique(merged_df$nitr)
merged_df$nitr <- factor(merged_df$nitr,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#pro
#unique(merged_df$pro)
merged_df$pro <- factor(merged_df$pro,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#diuretic
#unique(merged_df$diuretic)
merged_df$diuretic <- factor(merged_df$diuretic,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#proto
#unique(merged_df$proto)
merged_df$proto <- factor(merged_df$proto,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#exang
#unique(merged_df$exang)
merged_df$exang <- factor(merged_df$exang,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#xhypo
#unique(merged_df$xhypo)
merged_df$xhypo <- factor(merged_df$xhypo,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#slope
#unique(merged_df$slope)
# check count of 0 valuyes
#sum(merged_df$slope == 0, na.rm = TRUE)
# replace 0 with NA
merged_df <- merged_df %>%
  mutate(slope = ifelse(slope == 0, NA, slope))
# factor
merged_df$slope <- factor(merged_df$slope,
                     levels = c(1, 2, 3),
                     labels = c("Upsloping", "Flat", "Downsloping"),
                     ordered = TRUE
                     )
#restwm
#unique(merged_df$restwm)
# check count of 0 valuyes
# factor
merged_df$restwm <- factor(merged_df$restwm,
                     levels = c(0, 1, 2, 3),
                     labels = c("None", "Mild/Moderate", "Moderate/Severe", 
                                "Akinesis"),
                     ordered = TRUE
                     )
#thal
#unique(merged_df$thal)
# examine 1, 2, 5 values which are not in the documentation
# check count of 1,2,5 values
#sum(merged_df$thal == 1, na.rm = TRUE)
#sum(merged_df$thal == 2, na.rm = TRUE)
#sum(merged_df$thal == 5, na.rm = TRUE)
# replace 1, 2, 5 with na
merged_df <- merged_df %>%
  mutate(thal = ifelse(thal == 1, NA, thal)) %>%
  mutate(thal = ifelse(thal == 2, NA, thal)) %>%
  mutate(thal = ifelse(thal == 5, NA, thal))
```


## Data Cleaning

This below section of code starts by removing columns that are unused as indicated by the hear-disease.names text file which serves as a data dictionary and source for information on all the data sets. Once the explicitly unnecessary columns are removed, we convert the -9's and -9.0's into NA values and set a threshold for feature removal of if the feature is missing more than 60% of the observations it is immediately dropped. Once these features have been removed we focus on removing near zero variance features.

```{r data_cleaning}
# remove unnecessary columns as indicated in heart-disease.names
# cols marked as not used or dummy variables
not_used_dummy_var <- c(
  "id", "ccf", "dummy", "restckm", "exerckm", "thalsev", "thalpul",
  "earlobe", "lvx1", "lvx2", "lvx3", "lvx4", "lvf", "cathef", "junk", "name"
)

df_cleaning <- merged_df %>%
  select(-all_of(not_used_dummy_var))

# convert -9 to NA
df_cleaning <- df_cleaning %>%
  mutate(across(where(is.character), ~ na_if(.x, "-9"))) %>%   # character columns
  mutate(across(where(is.numeric), ~ na_if(.x, -9)))           # numeric columns
#Get NA count
sum(is.na(df_cleaning))
#get dimensions of the data
df_cleaning %>% dim
#reexamine the head
df_cleaning %>% head()

# calculate percentage missing
missing_threshold <- 60
missing_pct <- df_cleaning %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_percent") %>%
  arrange(desc(missing_percent))

# filter by col missing
missing_cols <- missing_pct %>%
  filter(missing_percent > missing_threshold)
var_missing_gt_60 <- missing_cols$column

missing_cols
# drop cols missing greater than threshold
df_cleaning <- df_cleaning %>%
  select(-all_of(var_missing_gt_60))
dim(df_cleaning)

# check near zero variance vars
nzv_cols <- nearZeroVar(df_cleaning %>% 
    select(-num), saveMetrics = TRUE)
nzv_cols %>% filter(nzv == TRUE)

# get column names of nzv
nzv_names <- nzv_cols %>%
  filter(nzv ==TRUE) %>%
  rownames()
nzv_names

# remove nzv variables
df_cleaning <- df_cleaning %>%
  select(-all_of(nzv_names))

# add back ca column for EDA
df_cleaning$ca <- merged_df$ca
df_cleaning$ca <- na_if(df_cleaning$ca, -9)

dim(df_cleaning)

# export to file
write.csv(df_cleaning, here("Data/Clean", "heart_disease_cleaned.csv"), row.names = FALSE)
saveRDS(df_cleaning, here("Data/Clean", "heart_disease_cleaned.rds"))
```

In the process of removing variables with >60% missing data, the `ca`
variable, which represents _Major vessels via fluoroscopy_ was removed
due to having 67.6% missing data. This variable was part of the 
_cleveland-14_ set. It has re-added to the final data set after _nzv_ variables
were removed. 

## Exploratory Data Analysis

```{r EDA}
df <- readRDS(here("Data/Clean", "heart_disease_cleaned.rds"))
# head(df)

missing_var <- df %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct")

missing_var %>%
  filter(missing_pct > 0) %>%
  arrange(desc(missing_pct))

# visualize missingness by variable
gg_miss_var(df)

# missingness patterns
library(VIM)
aggr(df, numbers = TRUE, sortVars = TRUE, cex.axis = 0.7)
# missingness correlation
gg_miss_upset(df, nsets = 10)
```

### Variables Missing > 40%
- ca: cleveland-14 variable - represents # of major vessels (0 - 3) colored by fluoroscopy.
- thal: cleveland-14 variable - results from thallium stress test
- thaltime: time duration for thallium stress test
- years: number of years as a smoker
- rldv5: height at rest
- famhist: family history of coronary artery disease
- cigs: cigarettes smoked per day

### Upset
- based on upset table, `ca`, `thal`, and `thaltime` are missing together
- missingness might not be random

```{r ca}
# create missingness columns for cleveland-14 variables
# histogram

show_histogram <- function(dataframe, feature) {
  ggplot(dataframe, aes(x = {{feature}})) + 
    geom_histogram(binwidth = 1, fill = "steelblue", color = "white") + 
  labs(
    title = paste("Histogram of", rlang::as_label(enquo(feature))), 
    x = rlang::as_label(enquo(feature)), 
    y = "Count"
  ) + 
  theme_minimal()
}


show_densityplot <- function(dataframe, feature) {
  ggplot(df, aes(x = {{ feature }})) +
    geom_density(fill = "darkorange", alpha = 0.5) +
    labs(
      title = paste("Density Plot of", rlang::as_label(enquo(feature))), 
      x = rlang::as_label(enquo(feature)), 
      y = "Count"
    ) +
    theme_minimal()
}

show_histogram(df, ca)
show_densityplot(df, ca)

# impute with median and missingness flag
df <- df %>%
  mutate(
    ca_missing = ifelse(is.na(ca), 1, 0),
    ca_imputed = ifelse(is.na(ca), median(ca, na.rm = TRUE), ca)
  )
```

```{r thal}
show_histogram(df, thal)
show_densityplot(df, thal)
table(df$thal)

# impute with mode and missingness flag
df <- df %>%
  mutate(
    thal_missing = ifelse(is.na(df$thal), 1, 0),
    thal_imputed <- ifelse(is.na(df$thal), 3, df$thal) # mode imputation
  )
```

```{r thaltime}
summary(df$thaltime)
sort(table(df$thaltime), decreasing = TRUE)

# histogram
show_histogram(df, thaltime)
show_densityplot(df, thaltime)

# impute with median + missingness flag
df <- df %>%
  mutate(
    thaltime_missing = ifelse(is.na(thaltime), 1, 0),
    thaltime_imputed = ifelse(is.na(thaltime), median(thaltime, na.rm = TRUE), thaltime)
  )
```

