---
title: "2_EDA"
output: pdf_document
date: "2025-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Cleaning and Exploratory Data Analysis

This R markdown document will be used to perform Exploratory Data Analysis (EDA) on the merged_df_raw that was developed in the previous workbook `1_data_extraction.Rmd`.

## Data Importing

```{r libraries}
library(tidyverse)
library(readr)
library(ggplot2)
library(corrplot)
library(caret)
library(here)
library(naniar)
library(VIM)
library(skimr)
library(stringr)
library(rcompanion)
```

```{r data_import}
merged_df <- read_csv(here('Data/Raw', 'merged_df_raw.csv')) 
head(merged_df)
```

## Categorical Conversions

```{r}
## categorical features and their levels and labels
#sex
merged_df$sex <- factor(merged_df$sex, 
                    levels = c(0, 1), 
                    labels = c("Female", "Male")
                    )
#painloc
#unique(merged_df$painloc)
merged_df$painloc <- factor(merged_df$painloc,
                        levels = c(0, 1), 
                        labels = c("Otherwise", "Substernal")
                        )
#painexer
#unique(merged_df$painexer)
merged_df$painexer <- factor(merged_df$painexer,
                        levels = c(0, 1), 
                        labels = c("Otherwise", "Provoked by Exertion")
                        )
#relrest
#unique(merged_df$relrest)
merged_df$relrest <- factor(merged_df$relrest,
                        levels = c(0, 1), 
                        labels = c("Otherwise", "Relieved after rest")
                        )
#cp
#unique(merged_df$cp)
merged_df$cp <- factor(merged_df$cp,
                        levels = c(1, 2, 3, 4), 
                        labels = c("Typical Angina", "Atypical Angina", 
                                   "Non-anginal Pain", "Asymptomatic"), 
                   ordered = TRUE)
#smoke
#unique(merged_df$smoke)
merged_df$smoke <- factor(merged_df$smoke,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#fbs
#unique(merged_df$fbs)
merged_df$fbs <- factor(merged_df$fbs,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#dm
#unique(merged_df$dm)
merged_df$dm <- factor(merged_df$dm,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#famhist
#unique(merged_df$famhist)
merged_df$famhist <- factor(merged_df$famhist,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#restecg
#unique(merged_df$restecg)
merged_df$restecg <- factor(merged_df$restecg,
                        levels = c(0, 1, 2), 
                        labels = c("Normal", "Abnormal", "Hypertrophy"), 
                   ordered = TRUE)
#dig
#unique(merged_df$dig)
merged_df$dig <- factor(merged_df$dig,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#prop
#unique(merged_df$prop)
# find the number of 22's a feature outside the range/
#sum(merged_df$prop == 22, na.rm = TRUE)
# convert the 22 to NA
merged_df <- merged_df %>%
  mutate(prop = ifelse(prop == 22, NA, prop))
#create factor
merged_df$prop <- factor(merged_df$prop,
                        levels = c(0, 1), 
                        labels = c("No", "Yes")
                        )
#nitr
#unique(merged_df$nitr)
merged_df$nitr <- factor(merged_df$nitr,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#pro
#unique(merged_df$pro)
merged_df$pro <- factor(merged_df$pro,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#diuretic
#unique(merged_df$diuretic)
merged_df$diuretic <- factor(merged_df$diuretic,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#proto
#unique(merged_df$proto)
merged_df$proto <- factor(merged_df$proto,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#exang
#unique(merged_df$exang)
merged_df$exang <- factor(merged_df$exang,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#xhypo
#unique(merged_df$xhypo)
merged_df$xhypo <- factor(merged_df$xhypo,
                     levels = c(0, 1),
                     labels = c("No", "Yes")
                     )
#slope
#unique(merged_df$slope)
# check count of 0 valuyes
#sum(merged_df$slope == 0, na.rm = TRUE)
# replace 0 with NA
merged_df <- merged_df %>%
  mutate(slope = ifelse(slope == 0, NA, slope))
# factor
merged_df$slope <- factor(merged_df$slope,
                     levels = c(1, 2, 3),
                     labels = c("Upsloping", "Flat", "Downsloping"),
                     ordered = TRUE
                     )
#restwm
#unique(merged_df$restwm)
# check count of 0 valuyes
# factor
merged_df$restwm <- factor(merged_df$restwm,
                     levels = c(0, 1, 2, 3),
                     labels = c("None", "Mild/Moderate", "Moderate/Severe", 
                                "Akinesis"),
                     ordered = TRUE
                     )
#thal
#unique(merged_df$thal)
# examine 1, 2, 5 values which are not in the documentation
# check count of 1,2,5 values
#sum(merged_df$thal == 1, na.rm = TRUE)
#sum(merged_df$thal == 2, na.rm = TRUE)
#sum(merged_df$thal == 5, na.rm = TRUE)
# replace 1, 2, 5 with na
merged_df <- merged_df %>%
  mutate(thal = ifelse(thal == 1, NA, thal)) %>%
  mutate(thal = ifelse(thal == 2, NA, thal)) %>%
  mutate(thal = ifelse(thal == 4, NA, thal)) %>%
  mutate(thal = ifelse(thal == 5, NA, thal))
```

## Data Cleaning

This below section of code starts by removing columns that are unused as indicated by the hear-disease.names text file which serves as a data dictionary and source for information on all the data sets. Once the explicitly unnecessary columns are removed, we convert the -9's and -9.0's into NA values and set a threshold for feature removal of if the feature is missing more than 60% of the observations it is immediately dropped. Once these features have been removed we focus on removing near zero variance features.

```{r data_cleaning}
# remove unnecessary columns as indicated in heart-disease.names
# cols marked as not used or dummy variables
not_used_dummy_var <- c(
  "id", "ccf", "dummy", "restckm", "exerckm", "thalsev", "thalpul",
  "earlobe", "lvx1", "lvx2", "lvx3", "lvx4", "lvf", "cathef", "junk", "name"
)

df_cleaning <- merged_df %>%
  select(-all_of(not_used_dummy_var))

# convert -9 to NA
df_cleaning <- df_cleaning %>%
  mutate(across(where(is.character), ~ na_if(.x, "-9"))) %>%   # character columns
  mutate(across(where(is.numeric), ~ na_if(.x, -9)))           # numeric columns
#Get NA count
sum(is.na(df_cleaning))
#get dimensions of the data
df_cleaning %>% dim
#reexamine the head
df_cleaning %>% head()

# calculate percentage missing
missing_threshold <- 60
missing_pct <- df_cleaning %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_percent") %>%
  arrange(desc(missing_percent))

# filter by col missing
missing_cols <- missing_pct %>%
  filter(missing_percent > missing_threshold)
var_missing_gt_60 <- missing_cols$column

missing_cols
# drop cols missing greater than threshold
df_cleaning <- df_cleaning %>%
  select(-all_of(var_missing_gt_60))
dim(df_cleaning)

# check near zero variance vars
nzv_cols <- nearZeroVar(df_cleaning %>% 
    select(-num), saveMetrics = TRUE)
nzv_cols %>% filter(nzv == TRUE)

# get column names of nzv
nzv_names <- nzv_cols %>%
  filter(nzv ==TRUE) %>%
  rownames()
nzv_names

# remove nzv variables
df_cleaning <- df_cleaning %>%
  select(-all_of(nzv_names))

# add back ca column for EDA
df_cleaning$ca <- merged_df$ca
df_cleaning$ca <- na_if(df_cleaning$ca, -9)

dim(df_cleaning)

# export to file
write.csv(df_cleaning, here("Data/Clean", "heart_disease_cleaned.csv"), row.names = FALSE)
saveRDS(df_cleaning, here("Data/Clean", "heart_disease_cleaned.rds"))
```

In the process of removing variables with \>60% missing data, the `ca` variable, which represents *Major vessels via fluoroscopy* was removed due to having 67.6% missing data. This variable was part of the *cleveland-14* set. It has re-added to the final data set after *nzv* variables were removed.

## Exploratory Data Analysis

```{r EDA}
df <- readRDS(here("Data/Clean", "heart_disease_cleaned.rds"))
# head(df)

missing_var <- df %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct")

missing_var %>%
  filter(missing_pct > 0) %>%
  arrange(desc(missing_pct))

# visualize missingness by variable
gg_miss_var(df)

# missingness patterns
aggr(df, numbers = TRUE, sortVars = TRUE, cex.axis = 0.7)
# missingness correlation
gg_miss_upset(df, nsets = 10)
```

##### Variables Missing \> 40%

-   ca: cleveland-14 variable - represents \# of major vessels (0 - 3) colored by fluoroscopy.
-   thal: cleveland-14 variable - results from thallium stress test
-   thaltime: time duration for thallium stress test
-   years: number of years as a smoker
-   rldv5: height at rest
-   famhist: family history of coronary artery disease
-   cigs: cigarettes smoked per day

##### Upset

-   based on upset table, `ca`, `thal`, and `thaltime` are missing together
-   missingness might not be random

```{r custom_functions}
# function for showing histogram
show_histogram <- function(dataframe, feature) {
  ggplot(dataframe, aes(x = {{feature}})) + 
    geom_histogram(binwidth = 1, fill = "steelblue", color = "white") + 
  labs(
    title = paste("Histogram of", rlang::as_label(enquo(feature))), 
    x = rlang::as_label(enquo(feature)), 
    y = "Count"
  ) + 
  theme_minimal()
}
# function for showing desnsity plot
show_densityplot <- function(dataframe, feature) {
  ggplot(df, aes(x = {{ feature }})) +
    geom_density(fill = "darkorange", alpha = 0.5) +
    labs(
      title = paste("Density Plot of", rlang::as_label(enquo(feature))), 
      x = rlang::as_label(enquo(feature)), 
      y = "Count"
    ) +
    theme_minimal()
}

show_barplot <- function(dataframe, feature) {
  ggplot(dataframe, aes(x = {{feature}})) +
  geom_bar(fill = "steelblue") +
  labs(
    title = paste("Bar Plot of", rlang::as_label(enquo(feature))), 
      x = rlang::as_label(enquo(feature)),  
    y = "Count") +
  theme_minimal()
}
```

#### ca

```{r ca}

show_histogram(df, ca)
show_densityplot(df, ca)

# impute with median and missingness flag
df <- df %>%
  mutate(
    ca_missing = ifelse(is.na(ca), 1, 0),
    ca_imputed = ifelse(is.na(ca), median(ca, na.rm = TRUE), ca)
  )

# drop imputed column
df <- df %>%
  select(-ca)
```

`ca` has a right skew therefore median was used for imputation.

#### thal

```{r thal}
show_histogram(df, thal)
show_densityplot(df, thal)
table(df$thal)

# impute with mode and missingness flag
df <- df %>%
  mutate(
    thal_missing = ifelse(is.na(df$thal), 1, 0),
    thal_imputed = ifelse(is.na(df$thal), 3, df$thal) # mode imputation
  )

# drop imputed column
df <- df %>%
  select(-thal)
```
#### thaltime

```{r thaltime}
summary(df$thaltime)
sort(table(df$thaltime), decreasing = TRUE)

# histogram
show_histogram(df, thaltime)
show_densityplot(df, thaltime)

# impute with median + missingness flag
df <- df %>%
  mutate(
    thaltime_missing = ifelse(is.na(thaltime), 1, 0),
    thaltime_imputed = ifelse(is.na(thaltime), median(thaltime, na.rm = TRUE), thaltime)
  )

# drop imputed column
df <- df %>%
  select(-thaltime)
```

`thaltime` showed a right skew. Imputed with median.

#### years + cigs

```{r}
show_histogram(df, years)
show_densityplot(df, years) 

show_histogram(df, cigs)
show_densityplot(df, cigs)

ggplot(df, aes(x = years, y = cigs)) + geom_point()
cor(df$years, df$cigs, use = "complete.obs", method = "pearson")

# create smoke_profile feature
# generative AI was used to create this code to create the smoke_profile feature
df <- df %>%
  mutate(
    # Categorize smoking duration
    years_cat = case_when(
      is.na(years)        ~ "missing",
      years == 0          ~ "non",
      years <= 10         ~ "short",
      years <= 20         ~ "medium",
      years <= 40         ~ "long",
      TRUE                ~ "very_long"
    ),

    # Categorize cigarette intensity
    cigs_cat = case_when(
      is.na(cigs)         ~ "missing",
      cigs == 0           ~ "non",
      cigs <= 10          ~ "light",
      cigs <= 20          ~ "moderate",
      cigs <= 40          ~ "heavy",
      TRUE                ~ "very_heavy"
    ),

    # Combine into single profile
    smoke_profile = case_when(
      years_cat == "non" & cigs_cat == "non"             ~ "non",
      years_cat == "missing" | cigs_cat == "missing"     ~ "missing",
      TRUE                                                ~ paste0(cigs_cat, "_", years_cat)
    ),

    # Optional: convert to factor
    smoke_profile = factor(smoke_profile)
  )

# drop helper columns
df <- df %>%
  select(-c(cigs_cat, years_cat))

# drop imputed column
df <- df %>%
  select(-c(years, cigs))
```

`years` and `cigs` has moderate colinearity. Due to missing data, a `smoke_profile` was engineered to combine the two columns.

#### rldv5

```{r rldv5}
show_histogram(df, rldv5)
show_densityplot(df, rldv5)

df <- df %>%
  mutate(
    rldv5_missing = ifelse(is.na(rldv5), 1, 0), 
    rldv5_imputed = ifelse(is.na(rldv5), median(rldv5, na.rm = TRUE), rldv5)
  )

# drop imputed column
df <- df %>%
  select(-rldv5)
```

`rldv5` has a right skew therefore imputed with median.

#### famhist

```{r famhist}
show_barplot(df, famhist)
unique(df$famhist)

# replace NA with "Missing" 
df$famhist <- as.character(df$famhist)
df$famhist[is.na(df$famhist)] <- "Missing"
df$famhist <- factor(df$famhist)

show_barplot(df, famhist)
```

Missing values in `famhist` were imputed with 'missing'.

### Drop original columns after imputation

```{r drop_imputed_gt40}
# list of cols
# missing_gt40 <- c('ca', 'thal', 'thaltime', 'years', 'rldv5', 'famhist', 'cigs')
imputed_cols <-c('ca_imputed', 'ca_missing', 'thal_imputed', 'thal_missing', 
                 'thaltime_missing', 'thaltime_imputed', 'smoke_profile', 
                 'rldv5_missing', 'rldv5_imputed', 'famhist')
# drop date / time columns
drop_cols <- c('cmo', 'cday', 'cyr', 'ekgmo', 'ekgday', 'ekgyr')
df1 <- df %>%
  select(-all_of(drop_cols))

# compare dimensions
dim(df_cleaning)
dim(df)
dim(df1)

colnames(df1)

head(df1)
```

## Missing < 40% 

### Binary Variables

```{r remaining_cols}
# list binary var
df1 %>% 
  summarise(across(everything(), ~ n_distinct(na.omit(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_unique") %>%
  filter(n_unique == 2)
# get binary var names
binary_vars <- df1 %>%
  select(where(~ n_distinct(na.omit(.)) == 2)) %>%
  names()

# convert binary vars to factors
df1[binary_vars] <- lapply(df1[binary_vars], as.factor)
# convert sex to male/female
#df$sex <- factor(df$sex, levels = c(0, 1), labels = c("female", "male"))

head(df1)
```
### Numerical Variables

```{r numerical_with_na}
numeric_vars <- df1 %>%
  select(where(is.numeric)) %>%
  names() 

numeric_na <- df1 %>%
  select(where(is.numeric)) %>%
  select(where(~ any(is.na(.)))) %>%
  names()

numeric_vars
df1[numeric_na]

# impute w median and create missingness columns
for (var in numeric_na) {
  median_val <- median(df1[[var]], na.rm = TRUE)
  df1[[paste0(var, "_missing")]] <- ifelse(is.na(df1[[var]]), 1, 0)
  df1[[paste0(var, "_imputed")]] <- ifelse(is.na(df1[[var]]), median_val, df1[[var]])
}

# drop imputed columns
df1 <- df1 %>%
  select(-all_of(numeric_na))
dim(df1)
```

Numerical variables were imputed with the median value. Missing flags were also created for these variables. 

### Categorical Variables

```{r categorical_with_na}

categorical_na <- df1 %>%
  select(where(~ is.factor(.) || is.character(.))) %>%
  select(where(~ any(is.na(.)))) %>%
  names()

# impute with Missing
for (var in categorical_na) {
  df1[[var]] <- as.character(df1[[var]])
  df1[[var]][is.na(df1[[var]])] <- "Missing"
  df1[[var]] <- factor(df1[[var]])
}
dim(df1)

anyNA(df1)
```
For NAs in categorical variables, NAs were replaced with 'Missing'. 

#### Create binary for target (num)

```{r}
# create binary target 0 = no disease, 1 - 3 = varying levels
df1 <- df1 %>%
  mutate(
    has_disease = factor(ifelse(num == 0, "No", "Yes"), levels = c("No", "Yes"))
  )

head(df1)
```
Binary target column for the target variable `num` was created. This column was left in. 

```{r export_imputed}
# export final dataset
saveRDS(df1, here("Data/Clean", "heart_disease_final.rds"))

```


# Univariate Analysis

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
df <- readRDS(here("Data/Clean", "heart_disease_final.rds"))
head(df)

# separate target from variables
target_vars <- c("num", "has_disease")
y_df <- df %>%
  select(all_of(target_vars))
         
X_df <- df %>%
  select(-all_of(target_vars))
```

## Numerical vars 

```{r}
num_vars <-X_df %>%
  select(where(is.numeric)) %>%
  names()

# summary 
X_df %>%
  select(all_of(num_vars)) %>%
  summary()


# Example: just the *_imputed columns (exclude *_missing)
imputed_vars <- names(X_df)[str_detect(names(X_df), "_imputed") & !str_detect(names(X_df), "_missing")]

X_df %>%
  select(all_of(imputed_vars)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "steelblue", outlier.shape = 16, outlier.size = 1) +
  theme_minimal() +
  labs(title = "Boxplots of Imputed Numeric Variables", x = "Variable", y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
```{r}
# summary statistics
X_df %>%
  select(where(is.numeric)) %>%
  skimr::skim()
```


## Categorical vars

```{r}
cat_vars <- X_df %>%
  select(where(~ is.factor(.) || is.character(.))) %>%
  names()

# relative frequency tables
for (var in cat_vars) {
  cat("----", var, "----\n")
  print(prop.table(table(X_df[[var]])))
  cat("\n")
}

```
```{r barplots-loop, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
# custom function for bar plots
plot_bar <- function(dataframe, var) {
  ggplot(dataframe, aes(x = .data[[var]])) +
    geom_bar(fill = "steelblue") +
    labs(title = paste("Bar Plot of", var), x = var, y = "Count") +
    theme_minimal()
}

for (var in cat_vars) {
  print(plot_bar(X_df, var))
}
```


## Feature Types

```{r}
X_df %>%
  summarise(
    numeric = sum(sapply(., is.numeric)),
    categorical = sum(sapply(., is.factor))
  )

```

# Bivariate Analysis

## Numerical Vars

### Box Plots

```{r}
plot_bivariate_num <- function(var) {
  ggplot(df, aes(x = factor(has_disease), y = .data[[var]])) +
    geom_boxplot(fill = "skyblue") +
    labs(title = paste(var, "by Disease Status"), x = "Has Disease", y = var) +
    theme_minimal()
}

for (var in num_vars){ 
print(plot_bivariate_num(var))
}

```

### Welch's two-sample t-test

```{r}
for (var in num_vars) {
  formula <- reformulate("has_disease", response = var)
  test_result <- t.test(formula, data = df)
  cat("=== T-test for", var, "===\n")
  print(test_result)
  cat("\n\n")
}
```

### Summary for numerical variables

```{r}
missing_flags <- colnames(df)[str_detect(colnames(df), "_missing$")]

# recommendation code generated with the help of ChatGPT. 
# Create an empty data frame to store results
bivariate_summary <- data.frame(
  Variable = character(),
  p_value = numeric(),
  Missingness_Flag = logical(),
  Recommendation = character(),
  stringsAsFactors = FALSE
)


# Loop through and compute t-tests
for (var in num_vars) {
  formula <- reformulate("has_disease", response = var)
  p_val <- tryCatch(t.test(formula, data = df)$p.value, error = function(e) NA)
  
  is_missing_flag <- paste0(var, "_missing") %in% colnames(df) | var %in% missing_flags

  recommendation <- if (!is.na(p_val) && p_val < 0.05) {
    "Include (predictive)"
  } else if (is_missing_flag) {
    "Include (missingness informative)"
  } else {
    "Optional (evaluate with model)"
  }
  
  bivariate_summary <- rbind(
    bivariate_summary,
    data.frame(
      Variable = var,
      p_value = round(p_val, 5),
      Missingness_Flag = is_missing_flag,
      Recommendation = recommendation
    )
  )
}

# View result
bivariate_summary

# create list of variables recommended as Include
selected_num_vars <- bivariate_summary %>%
  filter(str_detect(Recommendation, "Include")) %>%
  pull(Variable)
```

Based on recommendation, all variables recommended as 'Include' will be kept.


## Categorical Vars

### Bar Plots
```{r}
plot_bivariate_cat <- function(var) {
  ggplot(df, aes(x = .data[[var]], fill = factor(has_disease))) +
    geom_bar(position = "fill") +  # proportion stacked
    labs(title = paste(var, "by Disease Status"), y = "Proportion", fill = "Disease") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

for (var in cat_vars) {
  print(plot_bivariate_cat(var))
}
```

Have similar distributions: "prop", "nitr", "pro", "diuretic", "exang"
Have similar distributions:  "ladprox", "laddist", "cxmain", "om1", "rcaprox", "rcadist"

```{r}
cat_vars_to_keep <- c("sex", "painexer", "relrest", "cp", "htn", "fbs", "restecg", "exang", "slope", "ladprox", "location", "ca_missing", "thatltime_missing", "smoke_profile")
```

```{r}
# cramer v
cramerv <- function(dataframe, var_list){
  cramerV_matrix <- matrix(NA, nrow = length(var_list), ncol = length(var_list),
                           dimnames = list(var_list, var_list))
  
  for (i in seq_along(var_list)) {
    for (j in seq_along(var_list)) {
      tbl <- table(dataframe[[var_list[i]]], dataframe[[var_list[j]]])
      cramerV_matrix[i, j] <- suppressWarnings(cramerV(tbl))
    }
  }
  print(cramerV_matrix)
}

# Check correlation of variables that to have similar distribution
cramerv(df, c("prop", "nitr", "pro", "diuretic", "exang"))
cramerv(df, c("ladprox", "laddist", "cxmain", "om1", "rcaprox", "rcadist"))

drop_colinearity_cols <- c("prop", "nitr", "pro", "diuretic", "laddist", 
                           "cxmain", "om1", "rcaprox", "rcadist")
```

Based on the results, each set of variables are moderate to highly correlated. 
If necessary, these variables could be kept `exang` and `ladprox` while dropping the others. 

### Summary for categorical variables

```{r}
# recommendation code generated with the help of ChatGPT. 
cat_summary <- data.frame(
  Variable = character(),
  p_value = numeric(),
  Recommendation = character(),
  stringsAsFactors = FALSE
)

for (var in cat_vars) {
  tbl <- table(df[[var]], df$has_disease)
  p_val <- tryCatch(chisq.test(tbl)$p.value, error = function(e) NA)

  rec <- if (!is.na(p_val) && p_val < 0.05) {
    "Include (predictive)"
  } else {
    "Optional (evaluate with model)"
  }

  cat_summary <- rbind(
    cat_summary,
    data.frame(Variable = var, p_value = round(p_val, 5), Recommendation = rec)
  )
}

cat_summary

# create list of all variables recommended as Include
selected_cat_vars <- cat_summary %>%
  filter(str_detect(Recommendation, "Include")) %>%
  pull(Variable)

selected_cat_vars
```


## Final Variable Sets

```{r}
# combine selected variables
included_vars <- c(selected_num_vars, selected_cat_vars)

model_df <- df %>%
  select(all_of(included_vars), has_disease) %>%
  select(-all_of(drop_colinearity_cols))

str(model_df)
```

```{r}
# export modeling dataset to RDS file
saveRDS(model_df, here("Data/Clean", "heart_disease_modeling.rds"))

model_df %>% dim()
```

The final modeling set has 899 observations and 39 features including the target variable. 