---
title: "3_EDA Univariate and Bivariate Analysis"
output: pdf_document
date: "2025-06-13"
---

# Univariate Analysis

```{r}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(readr)
library(ggplot2)
library(corrplot)
library(caret)
library(here)
library(skimr)
library(stringr)
library(rcompanion)
```

```{r}
df <- readRDS(here("Data/Clean", "heart_disease_final.rds"))
head(df)

# separate target from variables
target_vars <- c("num", "has_disease")
y_df <- df %>%
  select(all_of(target_vars))
         
X_df <- df %>%
  select(-all_of(target_vars))
```

## Numerical vars 

```{r}
num_vars <-X_df %>%
  select(where(is.numeric)) %>%
  names()

# summary 
X_df %>%
  select(all_of(num_vars)) %>%
  summary()


# Example: just the *_imputed columns (exclude *_missing)
imputed_vars <- names(X_df)[str_detect(names(X_df), "_imputed") & !str_detect(names(X_df), "_missing")]

X_df %>%
  select(all_of(imputed_vars)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "steelblue", outlier.shape = 16, outlier.size = 1) +
  theme_minimal() +
  labs(title = "Boxplots of Imputed Numeric Variables", x = "Variable", y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
```{r}
# summary statistics
X_df %>%
  select(where(is.numeric)) %>%
  skimr::skim()
```


## Categorical vars

```{r}
cat_vars <- X_df %>%
  select(where(~ is.factor(.) || is.character(.))) %>%
  names()

# relative frequency tables
for (var in cat_vars) {
  cat("----", var, "----\n")
  print(prop.table(table(X_df[[var]])))
  cat("\n")
}

```
```{r barplots-loop, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
# custom function for bar plots
plot_bar <- function(dataframe, var) {
  ggplot(dataframe, aes(x = .data[[var]])) +
    geom_bar(fill = "steelblue") +
    labs(title = paste("Bar Plot of", var), x = var, y = "Count") +
    theme_minimal()
}

for (var in cat_vars) {
  print(plot_bar(X_df, var))
}
```


## Feature Types

```{r}
X_df %>%
  summarise(
    numeric = sum(sapply(., is.numeric)),
    categorical = sum(sapply(., is.factor))
  )

```

# Bivariate Analysis

## Numerical Vars

### Box Plots

```{r}
plot_bivariate_num <- function(var) {
  ggplot(df, aes(x = factor(has_disease), y = .data[[var]])) +
    geom_boxplot(fill = "skyblue") +
    labs(title = paste(var, "by Disease Status"), x = "Has Disease", y = var) +
    theme_minimal()
}

for (var in num_vars){ 
print(plot_bivariate_num(var))
}

```

### Welch's two-sample t-test

```{r}
for (var in num_vars) {
  formula <- reformulate("has_disease", response = var)
  test_result <- t.test(formula, data = df)
  cat("=== T-test for", var, "===\n")
  print(test_result)
  cat("\n\n")
}
```


### Summary for numerical variables

```{r}
missing_flags <- colnames(df)[str_detect(colnames(df), "_missing$")]

# recommendation code generated with the help of ChatGPT. 
# Create an empty data frame to store results
bivariate_summary <- data.frame(
  Variable = character(),
  p_value = numeric(),
  Missingness_Flag = logical(),
  Recommendation = character(),
  stringsAsFactors = FALSE
)


# Loop through and compute t-tests
for (var in num_vars) {
  formula <- reformulate("has_disease", response = var)
  p_val <- tryCatch(t.test(formula, data = df)$p.value, error = function(e) NA)
  
  is_missing_flag <- paste0(var, "_missing") %in% colnames(df) | var %in% missing_flags

  recommendation <- if (!is.na(p_val) && p_val < 0.05) {
    "Include (predictive)"
  } else if (is_missing_flag) {
    "Include (missingness informative)"
  } else {
    "Optional (evaluate with model)"
  }
  
  bivariate_summary <- rbind(
    bivariate_summary,
    data.frame(
      Variable = var,
      p_value = round(p_val, 5),
      Missingness_Flag = is_missing_flag,
      Recommendation = recommendation
    )
  )
}

# View result
bivariate_summary

# create list of variables recommended as Include
selected_num_vars <- bivariate_summary %>%
  filter(str_detect(Recommendation, "Include")) %>%
  pull(Variable)
```

Based on recommendation, all variables recommended as 'Include' will be kept.


## Categorical Vars

### Bar Plots
```{r}
plot_bivariate_cat <- function(var) {
  ggplot(df, aes(x = .data[[var]], fill = factor(has_disease))) +
    geom_bar(position = "fill") +  # proportion stacked
    labs(title = paste(var, "by Disease Status"), y = "Proportion", fill = "Disease") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

for (var in cat_vars) {
  print(plot_bivariate_cat(var))
}
```

Have similar distributions: "prop", "nitr", "pro", "diuretic", "exang"
Have similar distributions:  "ladprox", "laddist", "cxmain", "om1", "rcaprox", "rcadist"

```{r}
cat_vars_to_keep <- c("sex", "painexer", "relrest", "cp", "htn", "fbs", "restecg", "exang", "slope", "ladprox", "location", "ca_missing", "thatltime_missing", "smoke_profile")
```

```{r}
# cramer v
cramerv <- function(dataframe, var_list){
  cramerV_matrix <- matrix(NA, nrow = length(var_list), ncol = length(var_list),
                           dimnames = list(var_list, var_list))
  
  for (i in seq_along(var_list)) {
    for (j in seq_along(var_list)) {
      tbl <- table(dataframe[[var_list[i]]], dataframe[[var_list[j]]])
      cramerV_matrix[i, j] <- suppressWarnings(cramerV(tbl))
    }
  }
  print(cramerV_matrix)
}

# Check correlation of variables that to have similar distribution
cramerv(df, c("prop", "nitr", "pro", "diuretic", "exang"))
cramerv(df, c("ladprox", "laddist", "cxmain", "om1", "rcaprox", "rcadist"))

drop_colinearity_cols <- c("prop", "nitr", "pro", "diuretic", "laddist", 
                           "cxmain", "om1", "rcaprox", "rcadist")
```

Based on the results, each set of variables are moderate to highly correlated. 
If necessary, these variables could be kept `exang` and `ladprox` while dropping the others. 

### Summary for categorical variables

```{r}
# recommendation code generated with the help of ChatGPT. 
cat_summary <- data.frame(
  Variable = character(),
  p_value = numeric(),
  Recommendation = character(),
  stringsAsFactors = FALSE
)

for (var in cat_vars) {
  tbl <- table(df[[var]], df$has_disease)
  p_val <- tryCatch(chisq.test(tbl)$p.value, error = function(e) NA)

  rec <- if (!is.na(p_val) && p_val < 0.05) {
    "Include (predictive)"
  } else {
    "Optional (evaluate with model)"
  }

  cat_summary <- rbind(
    cat_summary,
    data.frame(Variable = var, p_value = round(p_val, 5), Recommendation = rec)
  )
}

cat_summary

# create list of all variables recommended as Include
selected_cat_vars <- cat_summary %>%
  filter(str_detect(Recommendation, "Include")) %>%
  pull(Variable)

selected_cat_vars
```


## Final Variable Sets

```{r}
# combine selected variables
included_vars <- c(selected_num_vars, selected_cat_vars)

model_df <- df %>%
  select(all_of(included_vars), has_disease) %>%
  select(-all_of(drop_colinearity_cols))

str(model_df)
```

```{r}
# export modeling dataset to RDS file
saveRDS(model_df, here("Data/Clean", "heart_disease_modeling.rds"))

model_df %>% dim()
```

The final modeling set has 899 observations and 39 features including the target variable. 